---
title: "1973-1974 Motor Trends Dataset Analysis"
author: "Richard Nacianceno"
date: "6/4/2020"
output: html_document
---

```{r Environment, message=FALSE}
library(tidyverse)
library(reshape2)
library(hablar)
library(RColorBrewer)
library(plotly)
library(ggbiplot) # installation library(devtools); install_github("vqv/ggbiplot")
sessionInfo()
```

## Dataset History

According to the help documentation ?mtcars: "The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models)."  

Description of variables:

1. mpg: Miles/(US) gallon
2. cyl: Number of cylinders
3. disp: Displacement (cu.in.)
4. hp: Gross horsepower
5. drat: Rear axle ratio
6. wt: Weight (1000 lbs)
7. qsec: 1/4 mile time
8. vs: 	Engine (0 = V-shaped, 1 = straight)
9. am: Transmission (0 = automatic, 1 = manual)
10. gear: Number of forward gears
11. carb: Number of carburetors

## Exploratory Data Analysis

### Formulate our question(s)

1. Is an automatic or manual transmission better for MPG?
2. Quantify the MPG difference between automatic and manual transmissions?

### Load Data
```{r load}
data(mtcars)
```

### Check the Packaging

Poke at the dataset. Ensure there are no errors on the given dataset. Check the top and bottom of the datasets.   
```{r head and tail}
head(mtcars)
tail(mtcars)
# view(mtcars)
```
Keep poking at it.  
```{r obs}
nrow(mtcars)
names(mtcars)
```

```{r variables}
ncol(mtcars)
rownames(mtcars)
```

### Run str()

This is a fast operation that provides us with the number of obersvations and variables, with their respective datatypes.  We can quickly analyze opportunities that we can see using str(). 
```{r datatype}
str(mtcars)
```
A few opportunities I see in the above output are as follows:

* The disp and hp are much higher compared to the remaining variables and rescaling could help with the analysis.  
* Variables cyl, vs, am, gear, carb are **potential** factor types

Using str() provides us with a quick and powerful way in analyzing our data and making it more tidy.

```{r change datatype}

str(mtcars)
```

### Missing Data Analysis

Analyze missing data and have a strategy in imputing missing datasets. 
```{r NA analysis}
naAnalysis <- mtcars %>%
    purrr::map_df(function(x) round(mean(is.na(x)),digits = 2)*100) %>%
    gather(EVType, naAverage)
naAnalysis %>% ggplot(aes(x = EVType, y = naAverage)) %>% +
  geom_point(aes(reorder(EVType, naAverage))) + theme(axis.text.x =
              element_text(angle = 90, hjust = .1)) + labs(x = "Event Type", 
              y = "NA Average  (%)", title = "Missing Data Analysis")
```

Based on our analysis there are no missing data in all our variables and no need to impute missing data.  

### Validate with at least one external data source

[According to the Buereau of Transportation](https://www.bts.gov/content/average-fuel-efficiency-us-light-duty-vehicles), the average fuel efficiency for ligh-duty vehicle-Passenger car in 2017 is 39.4 mpg.  

Also, we added another variable called countryGroup in mtcars which shows the country or continent they were made (personal assumption).

```{r mpg summary}
countryGroup <-c(rep("JAPAN", 3), rep("USA",4), rep("EUROPE", 7),rep("USA",3), "EUROPE", rep("JAPAN", 3), rep("USA",4), rep("EUROPE", 3), "USA", rep("EUROPE", 3))
mtcars$country <- countryGroup
quantile(mtcars$mpg, seq(0,1,0.1))
```

Above table, we zoomed-in our data using quantile function. If we are to add external dataset of **39.4 mpg** in mtcars dataset, this becomes the new max and a complete outlier. 

There are many technological advances and changes in environmental policies that occurred between 1973 to 2017. Some advancements are lighter metals and safety equipment to increase fuel efficiency and decrease risk in accident.  

There are many more advances in the field of vehicles-electrification that increases mpg-e(equivalent) and driverless vehicles (off-topic).

After 44 years, the advancement in fuel efficiency increased at an average rate of .44 mpg each year for 44 years or 19.31/44 years (note: 39.4-20.09=19.31). **This is pretty bad.** 

Tesla Model S on average has 115 city /107 highway mpg-e.  

Knowing the national average-mpg, gives us insight to the following:

* Right magnitude (mpg and not kilometers per gallon)
* The max MPG in our data set is at 33.90 vs. the mean of 39.4 mpg in 2017.
* The max outlier in year 1973 is less than the mean in 2017. 
* Adding the country of origin will provide us a better insight how these are group together.

Continue on the EDA...

### Principle Component Analysis

Let's review the relationship of variables to our vehicles:

```{r Principle Componnet Analysis}
pca <- prcomp(mtcars[, c(1:7,10,11)],center = TRUE, scale. = TRUE)
pca_x_df<- data.frame(pca[["x"]])
pca_name <- colnames(pca_x_df)
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca_df <- data.frame(name=pca_name,pca_percent =pca.var.per)
pca_df %>% ggplot(aes(x=name, y=pca_percent)) + geom_col()
```

```{r PCA Summary}
summary(pca)
```

```{r Plot PCA}
ggplotly(ggbiplot(pca, labels=rownames(mtcars), ellipse = TRUE, groups = mtcars$country))
```
Not all cars are made equal.  We can see from the above graph that the Japan, and Europe made vehicles tend to have a higher mpg and less hp, carb, cyl, disp, and wt compared with the USA counterpart. 

```{r pca_plot}
ggplotly(ggbiplot(pca, labels=rownames(mtcars), ellipse = TRUE, groups = mtcars$country, choices = c(3,4)))
```

  

Let's take a look at the breakdown of countries vs. am:

```{r }
group_country_am <- mtcars %>% group_by(country, am)
group_country_am <- dplyr::summarise(group_country_am, count=n())
group_country_am <-melt(group_country_am, id.vars = c("country", "am"))
ggplot(group_country_am, aes(x=country, y=value)) +
    geom_bar(stat="identity", alpha=.9)+facet_grid(~am)+labs(title = "Country by Auto Transmission (0=AT, 1=MT)", y= "Count", x= "Country/Continent")
```


```{r}
mpg_rank <- mtcars %>% arrange(desc(mpg)) %>% select(country, mpg, am) %>% mutate(am_abbr = ifelse(mtcars$am == 0, "AT", "MT"))

mpg_rank %>% mutate(ranking = rank(desc(mpg),ties.method = "first"))
```

Japan has a total of 5 vehicles in our sample size and the above chart shows that 4 of Japan vehicles are part of the top 10.  

```{r}
mtcars <- mtcars %>% convert(fct(cyl,vs, am, gear, carb))
```

### Try a simple solution

**Is an automatic or manual transmission better for MPG?**

Data set description of "am", from ?mtcars: 

* 0 = automatic transmission
* 1 = manual transmission
```{r Mean of Each Transmission}
amMean <-summarise(group_by(mtcars, AM=as.factor(am)),MN=mean(mpg))
mpgMean <- mean(mtcars$mpg)
amMean
```

```{r summary of each}
by(mtcars$mpg, INDICES = mtcars$am, summary)
```
**Manual transmission** has a higher min and max compared to **Automatic Transmission** in our dataset. Let's create a boxplot to visualize our findings.  

```{r boxplot, message=FALSE, error=FALSE}

g<-mtcars %>% ggplot(aes(y=mpg, x=am))
g+geom_boxplot(alpha=.3)+
  facet_grid(~mtcars$am)+
  geom_hline(aes(yintercept = mean(mpg), col="red"))+
  theme(legend.position = "none")+stat_summary(fun = mean, geom = "point",
  colour = "blue")
```
Transmission Analysis:

* Red line across boxplot is the total mpg mean, **`r mpgMean`**
* Blue point is our respective mean for AT and MT.  
* 0 or automatic transmission mean = **`r amMean[1,2]`**
* 1 or manual transmission mean = **`r amMean[2,2]`**
* The mean of total mpg is greater than Automatic Transmission mean or **`r mpgMean` >  `r amMean[1,2]`**
* The mean of the manual transmission is greater than the mean of the total mpg and is greater than the automatic transmission mpg or **`r amMean[2,2]` > `r mpgMean` > `r amMean[1,2]`**.

Let's review the counts of each transmission:

```{r Count of Transmission}
by_am <- mtcars %>% group_by(am)
dplyr::summarise(by_am, count = n())
```
```{r difference, echo=FALSE, results='hide'}
am_diff <-amMean[2,2]-amMean[1,2]
am_diff
```

**Answer to question 1:**

Based on the simple solution we applied, manual transmission is better than automatic transmission in fuel efficiency. We still need to keep poking our data for insights.  

**Quantify the MPG difference between automatic and manual transmissions?**

Manual Transmission mpg is `r am_diff` mpg higher than Automatic manual transmission mpg.  We do not know what percentage belongs to AM in the difference between MT and AT.  

We need to dig deeper in our analysis and keep poking our simple solution. 

### Deep Dive and Challenge "simple solution" and Build a Model

Hypothesis:

H0: AT mpg = MT mpg; There is no difference in fuel efficiency between Automatic transmission and Manual Transmission

Ha: AT mpg != MT mpg

```{r t.test}
t.test(mtcars$mpg~factor(mtcars$am), paired = FALSE, var.equal = TRUE)
```
There is a .0285% chance that our null hypothesis = MT mpg = MT mpg, therefore, we reject the null hypothesis. The mean of AT != mean of MT. 
Let's check out the linear model:

```{r linear model}
fit <- lm(data = mtcars, mpg~factor(am))
summary(fit)
```

As we said before we rejected the null hypothesis, the linear model shows the following:

1. Automatic Transmission has a mean of 17.147
2. The 7.245 is the estimated change in the intercept of the linear relatiohship between Manual Transmission mpg1 and Automatic Transmission mpg0, going from Automatic to Manual Transmission.  Also, `r fit$coef[1]` + `r fit$coef[2]` = the mean(Manual Transmission) = `r fit$coef[1] + fit$coef[2]`
3. The adjusted r-squared `r fit$ which means that our model only explains 33.8% of the variance. 

Let us review the plot and see if it is a good fit.

```{r poking at the simple solution, message=FALSE}
rm(list=ls())
attach(mtcars)
fit <- lm(am~mpg)
plot(mpg,am, frame.plot = FALSE, xlab = "MPG", ylab = "Transmission type (0=AT, 1=MT)")
abline(fit, col="blue")
lines(lowess(mpg, am), type="l", col="red")
```

The above linear model is not a good fit the blue bar crosses below 0 and above 1 and the redline crosses above 1. We need to find a better model that fits out data.

Log regression part of the GLM:

```{r finding the right fit, message=FALSE}
attach(mtcars)
fit <- glm(factor(mtcars$am)~mtcars$mpg, family = binomial)
plot(x=mpg,y=am, frame.plot = FALSE, xlab = "MPG", ylab = "Transmission type (0=AT, 1=MT)")
points(mpg,fit$fitted.values , col="red")
```

Exponentiated coefficient.
```{r exponentiated slope}
exp(fit$coefficients)
```

The first line shows the exponentiated slope coefficient is 1.36. We estimate a 36% increase in odds that the transmission is manual transmission per 1 mpg increase.  

```{r}
fit <- glm(factor(mtcars$am)~mtcars$mpg, family = binomial)
summary(fit)
```

The above model has an AIC of `r fit$aic` similar to the R-squared. Which means our model only accounts for 33.675% of the deviance. We need a better model.   

### Model Selection

Nested Model. 

1. mpg: Miles/(US) gallon
2. cyl: Number of cylinders
3. disp: Displacement (cu.in.)
4. hp: Gross horsepower
5. drat: Rear axle ratio
6. wt: Weight (1000 lbs)
7. qsec: 1/4 mile time
8. vs: 	Engine (0 = V-shaped, 1 = straight)
9. am: Transmission (0 = automatic, 1 = manual)
10. gear: Number of forward gears
11. carb: Number of carburetors

Let's scale down are data
```{r}
scale_mtcars <- data.frame(scale(mtcars))
fit1 <- lm(data = scale_mtcars, mpg~factor(am))
fit2 <- update(fit1, mpg~factor(am)+disp+hp)
fit3 <- update(fit1, mpg~factor(am)+disp+hp+drat+wt)
fit4 <- update(fit1, mpg~factor(am)+disp+hp+drat+wt+qsec+factor(vs))
fit5 <- update(fit1, mpg~factor(am)+disp+hp+drat+wt+qsec+factor(vs)+factor(cyl)+factor(gear)+factor(carb))
anova(fit1, fit2, fit3, fit4, fit5)
```

Model 2 and Model 3 seems to be a good candidate. Let us look deeper on model three and see if there is something we are missing.  

```{r}
fit1 <- lm(data = mtcars, mpg~factor(am))
fit2 <- update(fit1, mpg~factor(am)+disp+hp)
fit3 <- update(fit1, mpg~factor(am)+disp+hp+wt)
fit4 <- update(fit1, mpg~factor(am)+disp+hp+drat+wt+qsec+factor(vs))
fit5 <- update(fit1, mpg~factor(am)+disp+hp+drat+wt+qsec+factor(vs)+factor(cyl)+factor(gear)+factor(carb))
anova(fit1, fit2, fit3, fit4, fit5)
```


